---
title: "Example ADAT Workflow Using SomaScan.db"
abstract: "A short workflow illustrating the use of SomaScan.db to annotate SomaScan data (as an ADAT)."
package: "`r BiocStyle::pkg_ver('SomaScan.db')`"
output: 
  BiocStyle::html_document:
      toc_float: true
vignette: >
  %\VignetteIndexEntry{Example ADAT Workflow Using SomaScan.db}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, crop = NULL)
library(BiocStyle)
```

```{r loading packages}
suppressPackageStartupMessages({
    library(dplyr)
    library(GO.db)
    library(SomaDataIO)
    library(SomaScan.db)
})
```


# Introduction

This vignette will illustrate how the `SomaScan.db` package can be used to 
extend the results of basic bioinformatic analyses performed on SomaScan data. 
In this vignette, we will use tools from `r CRANpkg("SomaDataIO")`, in 
conjunction with `SomaScan.db`, to identify and annotate proteins 
(and ultimately, genes and their biological pathways) that are associated 
with a clinical variable of interest; e.g. `Age`. 

This workflow assumes the user has a basic knowledge of the R programming 
language. In addition, it is assumed that the user is familiar with the ADAT 
format, as well as the documentation in `r CRANpkg("SomaDataIO")` and the 
`r Githubpkg("SomaLogic/SomaLogic-Data")` Github repository. For more 
information about the ADAT format, as well as a detailed description of the 
fields found in an ADAT, please reference the 
`r Githubpkg("SomaLogic/SomaLogic-Data")` repository [README](https://github.com/SomaLogic/SomaLogic-Data).


# Reading an ADAT

The ADAT file format is a SomaLogic-specific, tab-delimited text file designed 
to store SomaScan study data. The example ADAT file used for this walkthrough, 
`example_data10.adat`, contains a SomaScan 5k study from a set of 10 human 
samples. The measurements and identifiers have been altered to protect 
personally identifiable information (PII), while retaining the underlying 
biological signal as much as possible. 

To begin, we will utilize `SomaDataIO::read_adat()` to load the example 
ADAT file:

```{r read-adat}
# Retrieve the example ADAT file from the SomaDataIO package
file <- system.file("extdata", "example_data10.adat",
                    package = "SomaDataIO", mustWork = TRUE
)
# Read in the file
adat <- SomaDataIO::read_adat(file)

adat
```

---------------------


## Data Prep and Exploration

Next, we will perform some preliminary data exploration of our variable 
of interest (`Age`):

```{r age-summary}
summary(adat$Age)
```

```{r age-hist, fig.height=3.5, fig.width=5}
hist(adat$Age,
     xlab = "Age (years)",
     main = "Age Distribution"
)
```

The subjects in this example data set are all adults, with age ranging from
`r min(adat$Age, na.rm = TRUE)` to `r max(adat$Age, na.rm = TRUE)` 
years.

However, as suggested by the summary table above, there is 1 patient that is 
missing age information:

```{r rm-na}
dplyr::filter(adat, is.na(Sex)) %>%
    dplyr::select(SlideId, Age)
```

**Note**: You may notice that the `select` function above has the namespace 
(`r CRANpkg("dplyr")`) specified; this is to distinguish between 
`dplyr::select()` and `SomaScan.db::select()`, both of which will be used in 
this vignette. A more thorough explanation of why this is important can be 
found in the "`select` namespace collision" section.

Samples corresponding to this patient will need to be removed before 
proceeding:

```{r}
adat <- dplyr::filter(adat, !is.na(Age))
```

As a quick sanity check: this ADAT was generated using the 5k SomaScan assay 
menu, which indicates that there should be RFU values for approximately 
5,000 analytes.

```{r n-analytes}
analytes <- SomaDataIO::getAnalytes(adat)

head(analytes)
```

```{r}
length(analytes)
```

Indeed, there are `r length(analytes)` analytes in this example ADAT file, 
consistent with the assay version.

---------------------


## Note on `SeqId` format

The values obtained in `analytes` above contains the same information as a 
SomaLogic `SeqId`, just converted to a more R-friendly format (i.e. the 
`seq`-prefix and `"."` delimiters). This format of identifier is 
used throughout `SomaDataIO`, and is also compatible with `SomaScan.db`. 
Values obtained from `SomaDataIO` (with the `seq`-prefix) are 
converted to `SeqIds` automatically when used by methods in `SomaScan.db`.


# Identifying Associations

We can now examine the data to see if any analytes are correlated with the 
variable `Age`. First, however, the data be be pre-processed. SomaLogic 
generally recommends to log-transform the RFU values in an ADAT prior to 
analysis.

```{r log-trans}
log10_adat <- log10(adat)
```

After log-transformation, we can examine the correlations and identify 
analytes that are positively correlated with `Age`. Here, correlations will be 
calculated using `stats::cor.test()`:

```{r cor-test}
# Calculate correlations for each analyte
cors_df <- lapply(analytes, function(x) {
    results <- stats::cor.test(log10_adat$Age, log10_adat[[x]],
                               method = "spearman", exact = FALSE)
    results <- results[c("estimate", "p.value")]
    unlist(results)
}) %>% setNames(analytes)

# Bind results together into a single dataframe
cors_df <- dplyr::bind_rows(cors_df, .id = "analytes")

# Isolate top positive correlations
top_posCor <- cors_df %>%
    dplyr::filter(p.value < 0.05) %>% # Retain significant cors only
    dplyr::filter(estimate.rho >= 0.75) %>% # Retain strong correlations
    arrange(desc(estimate.rho))

nrow(top_posCor)
head(top_posCor, 20L)
```

The table above contains analytes that have a strong positive 
correlation with `Age`, but this information alone may not enough to derive 
meaningful biological insights. Which proteins and genes do these identifiers 
correspond to? Are the most correlated proteins functionally related in some 
way, perhaps as part of the same biological pathway? These are the types of 
questions that `SomaScan.db` is designed to address.

In the next section, we will annotate these data by querying the above 
`SeqIds` in `SomaScan.db` to retrieve additional information about their 
corresponding genes and gene types.


# Annotating Results

To obtain a list of available annotations, use the `columns` method:

```{r columns}
columns(SomaScan.db)
```

Note that the "PROBEID" column corresponds to the SomaScan `SeqId`, the 
central probe for the platform. This identifier ties the other available 
annotations (listed in the `columns` output above) to the data found in an 
ADAT file. For more information on the ADAT file format and an explanation of 
its contents, please reference the [SomaLogic-Data GitHub repository](https://github.com/SomaLogic/SomaLogic-Data).

For this example, we will retrieve Gene Ontology (GO) identifiers associated 
with the `SeqIds` that are positively correlated with `Age`. This will be the 
first step in identifying biological processes associated with the protein 
targets of these `SeqIds`.

```{r GO-columns}
grep("GO|ONTOLOGY", columns(SomaScan.db), value = TRUE)
```

The `GO` and `GOALL` columns both contain GO pathway identifiers, while  
`ONTOLOGY` and `ONTOLOGYALL` contain additional metadata about the 
identifiers. For additional information about the values returned by `columns`,
run `help("{COLUMN NAME}")` to get a more detailed description of the listed 
options. 

---------------------


## Retrieving Annotations

The `select` method can be used to query the annotations to retrieve 
information corresponding to SomaScan analytes of interest. But, before we 
proceed, a note about the `select` method.

---------------------


### `select` Namespace Collisions

The `select` method is unique to Bioconductor's annotation packages (of which 
`SomaScan.db` is one), and should not be confused with the similarly-named 
`dplyr::select()`. These two functions perform very different actions on 
different classes of objects. If you have the `r CRANpkg("dplyr")` package 
([dplyr](https://dplyr.tidyverse.org)) loaded at the same time as 
`SomaScan.db` (as we do in this vignette), you may encounter an error like the 
one below when using `select`:

```r
Error in UseMethod("select") : 
  no applicable method for 'select' applied to an object of class "c('SomaDb', 
  'ChipDb', 'AnnotationDb', 'envRefClass', '.environment', 'refClass', 
  'environment', 'refObject', 'AssayData')"
```

This error indicates that it is unclear which `select` you are attempting to 
use (the `dplyr` version or the `SomaScan.db` version). To remedy this, it 
can be helpful to use the `::` operator to specify the package namespace 
when calling `select`, as seen in the code chunk below.

---------------------

### Gene Name Annotation

Continuing on with the example, we will retrieve gene names (as symbols) that 
correspond to the top 10 analytes associated with `Age`.

```{r top-10}
top10_analytes <- head(top_posCor$analytes, 10L)

anno <- SomaScan.db::select(SomaScan.db, 
                            keys = top10_analytes,
                            columns = c("SYMBOL", "GENENAME", "GENETYPE"))

anno
```

Remember that `select` retains the order of the original query, so these genes 
are still listed in order of most highly associated with `Age`. 

---------------------


### GO Annotation

Which pathways or biological processes are these genes associated with? Could 
that information help explain why they positively correlated with age? 
`SomaScan.db` enables mapping between `SeqIds` and GO identifiers, so let's 
take the top 3 genes and add GO annotations to the data frame. But why only 3 
genes? We are likely to receive a lot of additional 
information if we query the database for GO identifiers, so it can be easier 
and cleaner to begin with a small example data set when retrieving additional 
annotations.

```{r}
go_anno <- SomaScan.db::select(SomaScan.db, 
                               keys = anno$PROBEID[1:3L],
                               columns = c("SYMBOL", "GENENAME", "GENETYPE", 
                                           "GO", "ONTOLOGY")) %>%
    dplyr::filter(ONTOLOGY == "BP")

go_anno
```

As expected when working with GO terms, there were quite a few rows retrieved 
by that `select` query (`r nrow(go_anno)` rows of information for only 3 
genes). Filtering the results to the biological process ("BP") ontology only 
can help reduce the number of GO identifiers that are returned in the query.

This leaves us with a list of GO identifiers, but those identifiers do not 
adequately explain much about the term itself. How can we get more 
"human-readable" information out of GO?

Luckily, this is easy to do with other annotation tools from Bioconductor. The 
`r Biocpkg("GO.db")` annotation package contains annotations describing the  
entire Gene Ontology knowledgebase, assembled using data directly from the 
[GO website](http://geneontology.org/). We can use the GO IDs identified by 
`SomaScan.db` to connect with `r Biocpkg("GO.db")` and retrieve more 
information about each GO ID.

```{r}
go_terms <- AnnotationDbi::select(GO.db, 
                                  keys = go_anno$GO, 
                                  columns = c("GOID", "TERM", "DEFINITION"))

go_terms
```

Now we have _much_ more information about each GO ID! Using these IDs, we can 
merge this information back into our `select` results from `SomaScan.db`:

```{r join-go}
final_df <- left_join(go_anno, go_terms, by = c("GO" = "GOID"))

final_df
```

The `SomaScan.db` package can be used to link to numerous other
annotation resources. For a more detailed description of how this can be done
with examples of what resources are available, see the Advanced Usage Examples 
vignette.

There is still room for further exploration and extension of these results; 
this workflow is meant to be an introduction to how `SomaScan.db` can be 
used to build upon and interpret information obtained from a SomaLogic ADAT.


# Session Info

```{r session-info}
sessionInfo()
```
